---
title: "Effects of Mental Health on Substance Use"
author: "Nikita Johnson"
format: 
  html:
    toc: true
editor: visual
---

## **Introduction and Literature Review**

In 2022, there was 1 suicide death every 11 minutes (CDC, 2024). This is a staggering number, and doesn’t even account for the amount of people with serious thoughts or plans of suicide. According to the National Survey on Drug Use and Health in 2020, 4.9% of adults had serious thoughts of suicide in the past year, while 1.3% had made a suicide plan and 0.5% actually attempted suicide (SAMHSA, 2021). Women are significantly more likely to report past suicide attempts than men (Kelly et al., 2020) and young adults aged 18-25 are the most common group to be experiencing serious thoughts of suicide, plans, or attempts (SAMHSA, 2021; Piscopo & Lipari, 2016). Gay and bisexual men are also 4 times as likely to attempt suicide throughout their lifetime compared to heterosexual men, and lesbian and bisexual women are 2 times as likely to do so compared to heterosexual women. Depression and alcohol or other substance use disorders are also 1.5 times higher risk for lesbian, gay and bisexual people compared to heterosexual people (King et al., 2008). 

Suicidality is often comorbid with many other conditions and in this project, we will focus on its comorbidity with major depressive disorder and substance use disorder and how these differ for sexual minorities compared to heterosexual people. Having both a substance use disorder and major depressive disorder is associated with even greater psychological distress than only having one or the other. People with comorbid MDD and alcohol dependence in the past year are at least 20 times more likely to have suicidal ideation than those without both conditions. (Currie et al., 2005). Similarly, the people with the highest levels of past month alcohol use tend to also have recent suicidal ideation (Kelly et al., 2020). Serious suicidal ideation has also been found to be higher for people who binge drank frequently in the past month, used illicit drugs or alcohol in the past year, had a substance use disorder in the past year, and had a major depressive episode in the past year (Piscopo & Lipari, 2016). Current cigarette smokers are also at a higher risk of suicidal ideation, plans, attempts, and death compared to nonsmokers (Poorolajal & Darvishi, 2016). Heavy and long-term cannabis consumption has also been shown to be associated with worse suicidal outcomes (Shamabadi et al., 2023). Individuals who are dependent on substances are almost 3 times as likely to have MDD and around 4 times as likely to have suicidal thoughts compared to those who are not dependent on substances (Currie et al., 2005). 

Among people 12 or older in 2020, 10.2% had an alcohol use disorder, 6.6% had at least one illicit drug use disorder and 5.1% had a marijuana use disorder. Having any kind of substance use disorder was highest for young adults aged 18-25. 8.4% of adults also reported having a past year major depressive disorder in 2020, with 6% of adults being severely impaired by a major depressive episode in the past year. Those with a comorbid mental illness and a substance use disorder made up 6.7% of adults in 2020 (SAMHSA, 2021).

## **Research Question**

My main research question for this project is “How does mental health influence substance use in lesbian, gay and bisexual (LGB) individuals compared to their non-LGB counterparts?” Since there is no question asking about if respondents are transgender, we will only be focusing on LGB rather than LGBT.

Studies have shown that LGB individuals are more likely to experience depression, anxiety, psychological distress, and suicidality. This research question seeks to understand how these mental health issues might influence behaviors like substance use, which are often used as coping mechanisms. Substance use, including alcohol, tobacco, and drugs, can be both a result of mental health struggles and a contributor to further mental health decline.

Understanding how mental health influences substance use specifically in LGB individuals is essential to uncover whether mental health challenges are more likely to lead to substance use among LGB individuals compared to non-LGB individuals.

## **Hypotheses**

1\. LGB individuals who have experienced psychological distress in the past month are more likely to engage in regular substance use (alcohol, cigarette, and marijuana) compared to non-LGB individuals.

```{r}
#| message: false
#| warning: false
library(haven)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggdag)
library(dagitty)
library(psych)
library(car)
library(gtsummary)
library(broom.helpers)
library(expss)
library(pscl)
library(pROC)
library(arrow)
```

```{r, fig.alt = "Directed acyclic graph showing the relationship between the independent variable- psychological distress and the dependent variable- regular substance use. This relationship is hypothesized to be moderated by sexual orientaton."}
dag1 <- dagitty(
  'dag {
"Major Depressive Episode" [pos="-0.501,-1.065"]
"Psychological Distress (IV)" [exposure,pos="-1.475,0.334"]
"Regular Substance Use (DV)" [outcome,pos="0.364,0.334"]
"Sexual Orientation (Moderator)" [pos="-0.508,0.004"]
Age [pos="-0.508,1.236"]
Sex [pos="-0.510,0.714"]
Suicidality [pos="-0.506,-0.481"]
"Major Depressive Episode" -> "Psychological Distress (IV)"
"Major Depressive Episode" -> "Regular Substance Use (DV)"
"Psychological Distress (IV)" -> "Regular Substance Use (DV)"
"Sexual Orientation (Moderator)" -> "Psychological Distress (IV)"
"Sexual Orientation (Moderator)" -> "Regular Substance Use (DV)"
Age -> "Psychological Distress (IV)"
Age -> "Regular Substance Use (DV)"
Sex -> "Psychological Distress (IV)"
Sex -> "Regular Substance Use (DV)"
Suicidality -> "Psychological Distress (IV)"
Suicidality -> "Regular Substance Use (DV)"
}')

plot(dag1)
```

This hypothesis aims to explore the relationship between psychological distress in the past month (IV) and regular substance use in the past month (DV), which is hypothesized to be moderated by sexual orientation. Potential confounding variables that will need to be controlled for during regression are whether or not the respondent has experienced a major depressive episode in the past year, whether or not the respondent has had suicidal thoughts, plans, or attempts in the past year, as well as sex and age.

2\. LGB individuals who report experiencing a past year major depressive episode will be more likely to engage in regular substance use (alcohol, cigarette, and marijuana) compared to non-LGB individuals.

```{r, fig.alt = "Directed acyclic graph showing the relationship between the independent variable- major depressive episode and the dependent variable- regular substance use. This relationship is hypothesized to be moderated by sexual orientaton."}
dag2 <- dagitty(
  'dag {
  "Major Depressive Episode (IV)" [exposure,pos="-1.475,0.334"]
"Psychological Distress" [pos="-0.501,-1.065"]
"Regular Substance Use (DV)" [outcome,pos="0.364,0.334"]
"Sexual Orientation (Moderator)" [pos="-0.508,0.004"]
Age [pos="-0.508,1.236"]
Sex [pos="-0.510,0.714"]
Suicidality [pos="-0.506,-0.481"]
"Major Depressive Episode (IV)" -> "Regular Substance Use (DV)"
"Psychological Distress" -> "Major Depressive Episode (IV)"
"Psychological Distress" -> "Regular Substance Use (DV)"
"Sexual Orientation (Moderator)" -> "Major Depressive Episode (IV)"
"Sexual Orientation (Moderator)" -> "Regular Substance Use (DV)"
Age -> "Major Depressive Episode (IV)"
Age -> "Regular Substance Use (DV)"
Sex -> "Major Depressive Episode (IV)"
Sex -> "Regular Substance Use (DV)"
Suicidality -> "Major Depressive Episode (IV)"
Suicidality -> "Regular Substance Use (DV)"}'
)

plot(dag2)
```

This hypothesis aims to explore the relationship between experiencing a major depressive episode in the past year (IV) and regular substance use in the past month (DV), which is hypothesized to be moderated by sexual orientation. Potential confounding variables that will need to be controlled for during regression are whether or not the respondent has experienced psychological distress in the past month, whether or not the respondent has had suicidal thoughts, plans, or attempts in the past year, as well as sex and age.

3\. LGB individuals who have experienced serious psychological distress in the past month will be more likely to have a higher frequency of binge drinking in the past month compared to non-LGB individuals.

```{r, fig.alt = "Directed acyclic graph showing the relationship between the independent variable- psychological distress and the dependent variable- regular binge drinking. This relationship is hypothesized to be moderated by sexual orientaton."}
dag3 <- dagitty(
  'dag {
"Major Depressive Episode" [pos="-0.501,-1.065"]
"Psychological Distress (IV)" [exposure,pos="-1.475,0.334"]
"Regular Binge Drinking (DV)" [outcome,pos="0.364,0.334"]
"Sexual Orientation (Moderator)" [pos="-0.508,0.004"]
Age [pos="-0.508,1.236"]
Sex [pos="-0.510,0.714"]
Suicidality [pos="-0.506,-0.481"]
"Major Depressive Episode" -> "Psychological Distress (IV)"
"Major Depressive Episode" -> "Regular Binge Drinking (DV)"
"Psychological Distress (IV)" -> "Regular Binge Drinking (DV)"
"Sexual Orientation (Moderator)" -> "Psychological Distress (IV)"
"Sexual Orientation (Moderator)" -> "Regular Binge Drinking (DV)"
Age -> "Psychological Distress (IV)"
Age -> "Regular Binge Drinking (DV)"
Sex -> "Psychological Distress (IV)"
Sex -> "Regular Binge Drinking (DV)"
Suicidality -> "Psychological Distress (IV)"
Suicidality -> "Regular Binge Drinking (DV)"
}
'
)

plot(dag3)


```

This hypothesis aims to explore the relationship between experiencing psychological distress in the past month (IV) and regular binge drinking in the past month (DV), which is hypothesized to be moderated by sexual orientation. Potential confounding variables that will need to be controlled for during regression are whether or not the respondent has experienced a major depressive episode in the past year, whether or not the respondent has had suicidal thoughts, plans, or attempts in the past year, as well as sex and age.

4\. LGB individuals who have had suicidal thoughts, plans, or attempts in the past year will be more likely to have a substance use disorder compared to non-LGB individuals.

```{r, fig.alt = "Directed acyclic graph showing the relationship between the independent variable- suicidality and the dependent variable- substance use disorder. This relationship is hypothesized to be moderated by sexual orientaton."}
dag4 <- dagitty(
  'dag {
"Major Depressive Episode" [pos="-0.623,-1.054"]
"Psychological Distress" [pos="-0.625,-0.800"]
"Sexual Orientation (Moderator)" [pos="-0.632,-0.417"]
"Substance Use Disorder (DV)" [outcome,pos="0.251,-0.110"]
"Suicidality (IV)" [exposure,pos="-1.431,-0.110"]
Age [pos="-0.638,0.543"]
Sex [pos="-0.636,0.247"]
"Major Depressive Episode" -> "Substance Use Disorder (DV)"
"Major Depressive Episode" -> "Suicidality (IV)"
"Psychological Distress" -> "Substance Use Disorder (DV)"
"Psychological Distress" -> "Suicidality (IV)"
"Sexual Orientation (Moderator)" -> "Substance Use Disorder (DV)"
"Sexual Orientation (Moderator)" -> "Suicidality (IV)"
"Suicidality (IV)" -> "Substance Use Disorder (DV)"
Age -> "Substance Use Disorder (DV)"
Age -> "Suicidality (IV)"
Sex -> "Substance Use Disorder (DV)"
Sex -> "Suicidality (IV)"
}
'
)

plot(dag4)


```

This hypothesis aims to explore the relationship between experiencing suicidal thoughts, plans, or attempts in the past year (IV) and having a substance use disorder in the past year (DV), which is hypothesized to be moderated by sexual orientation. Potential confounding variables that will need to be controlled for during regression are whether or not the respondent has experienced a major depressive episode in the past year, whether or not the respondent has experienced psychological distress in the past month, as well as sex and age.

These hypotheses seem to be supported by previous literature in that people who have the poorest health outcomes tend to engage in regular substance use, binge drink more frequently, or have a substance use disorder. This project aims to expand on previous literature in the field by adding sexual orientation as a new lens to view these relationships with.

## **Data Cleaning and Descriptive Statistics**

The data found in this project is from the [2023 National Survey on Drug Use and Health](https://www.samhsa.gov/data/data-we-collect/nsduh-national-survey-drug-use-and-health/datafiles). This survey measures substance use, mental illness, and treatment in the civilian non-institutionalized population aged 12 or older. For the purposes of this project, however, I will only be focusing on adults. First, I will be cleaning the data and then I’ll introduce all the variables I plan to use in this project, along with a brief description of how the corresponding data looks in terms of descriptive statistics.

### **Data Cleaning**

First, we need to read in the data and keep only certain columns of interest. We will also rename the columns to make them more readable/understandable.

```{r}
path = file.path("~/GitHub/Effects/original_nsduh_data/NSDUH_2023.sav")

df = read_sav(path, col_select = c(CATAG6, IRSEX, CIG30USE, PMNICDEP, ALCDAYS, ALCBNG30D, MJDAY30A, IRPYUD5ALC, IRPYUD5MRJ, SPDPSTMON, ADSUITPAYR, AMDEYR, SEXIDENT22))

colnames(df) <- c('sex_orient', 'sex', 'age', 'cig30use', 'nicdepend', 'alc30use', 'alcbinge30', 'mj30use', 'alcdisorder_year', 'mjdisorder_year', 'psych_distress_30', 'suicide', 'majordepressive_year')

head(df)
```

Next, we will remove any observations from children since we want to focus on adults. We will also remove observations where people responded “don’t know”, “refused”, or left the answer blank. This will be done by going through each individual column and removing observations corresponding with certain values.

```{r}
#sex_orient
#responses with 6, 85, 94, 97, ad 98 will be removed
new_df <- subset(df, !(sex_orient %in% c(6, 85, 94, 97, 98)))

sex_orient_freq_table <- new_df %>%   
  group_by(sex_orient) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))  

sex_orient_freq_table
```

```{r}
#sex
sex_freq_table <- df %>%   
  group_by(sex) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))  

sex_freq_table
```

```{r}
#age
#responses with 1 will be removed to remove minors from dataset
new_df <- subset(df, age != 1)

age_freq_table <- new_df %>%   
  group_by(age) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))  

age_freq_table
```

```{r}
#cig30use
#responses with 94, 97, and 98 will be removed
new_df <- subset(new_df, !(cig30use %in% c(94, 97, 98)))

#responses with 91 and 93 will be changed to 0
new_df$cig30use[new_df$cig30use == 91] <- 0
new_df$cig30use[new_df$cig30use == 93] <- 0

cig30use_freq_table <- new_df %>%   
  group_by(cig30use) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))

cig30use_freq_table
```

```{r}
#nicdepend
#responses with NA will be removed
new_df <- new_df %>%
  drop_na(nicdepend)

#responses with 9 will be changed to 0
new_df$nicdepend[new_df$nicdepend == 9] <- 0

nicdepend_freq_table <- new_df %>%   
  group_by(nicdepend) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))

nicdepend_freq_table
```

```{r}
#alc30use
#responses with 85, 94, 97, and 98 will be removed
new_df <- subset(new_df, !(alc30use %in% c(85, 94, 97, 98)))

#responses with 91 and 93 will be changed to 0
new_df$alc30use[new_df$alc30use == 91] <- 0
new_df$alc30use[new_df$alc30use == 93] <- 0

alc30use_freq_table <- new_df %>%   
  group_by(alc30use) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))

alc30use_freq_table
```

```{r}
#alcbinge30
#responses with 85, 94, 97, and 98 will be removed
new_df <- subset(new_df, !(alcbinge30 %in% c(85, 94, 97, 98)))

#responses with 80, 91, and 93 will be changed to 0
new_df$alcbinge30[new_df$alcbinge30 == 80] <- 0
new_df$alcbinge30[new_df$alcbinge30 == 91] <- 0
new_df$alcbinge30[new_df$alcbinge30 == 93] <- 0

alcbinge30_freq_table <- new_df %>%   
  group_by(alcbinge30) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))

alcbinge30_freq_table
```

```{r}
#mj30use
#responses with 94, 97, and 98 will be removed
new_df <- subset(new_df, !(mj30use %in% c(94, 97, 98)))

#responses with 91 and 93 will be changed to 0
new_df$mj30use[new_df$mj30use == 91] <- 0
new_df$mj30use[new_df$mj30use == 93] <- 0

mj30use_freq_table <- new_df %>%   
  group_by(mj30use) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))

mj30use_freq_table
```

```{r}
#alcdisorder_year
alcdisorder_year_freq_table <- new_df %>%   
  group_by(alcdisorder_year) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))

alcdisorder_year_freq_table
```

```{r}
#mjdisorder_year
mjdisorder_year_freq_table <- new_df %>%   
  group_by(mjdisorder_year) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))

mjdisorder_year_freq_table
```

```{r}
#psych_distress_30
psych_distress_30_freq_table <- new_df %>%   
  group_by(psych_distress_30) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))

psych_distress_30_freq_table
```

```{r}
#suicide
#responses with 2 will be changed to 0
new_df$suicide[new_df$suicide == 2] <- 0

suicide_freq_table <- new_df %>%   
  group_by(suicide) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))

suicide_freq_table
```

```{r}
#majordepressive_year
#responses with 2 will be changed to 0
new_df$majordepressive_year[new_df$majordepressive_year == 2] <- 0

#responses with NA will be removed
new_df <- new_df %>%
  drop_na(majordepressive_year)

majordepressive_year_freq_table <- new_df %>%   
  group_by(majordepressive_year) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))

majordepressive_year_freq_table
```

Next, I will create new columns with the values of categorical variables for easier plotting later on.

```{r}
#sex_orient2
new_df <- new_df %>% 
  mutate(sex_orient2 = case_when(sex_orient == 1 ~ "Heterosexual",
                                 sex_orient == 2 ~ "Gay/Lesbian",
                                 sex_orient == 3 ~ "Bisexual",
                                 sex_orient == 4 ~ "Other",
                                 sex_orient == 5 ~ "Questioning"))
```

```{r}
#sex_orient3
#change sex_orient to be 0 = non-LGB and 1 = LGB
new_df <- new_df %>% 
  mutate(sex_orient3 = case_when(sex_orient == 1 ~ 0,
                                 sex_orient == 2 ~ 1,
                                 sex_orient == 3 ~ 1,
                                 sex_orient == 4 ~ 1,
                                 sex_orient == 5 ~ 1))
```

```{r}
#there are NA values in our new sex_orient2 column so we will remove those
new_df <- na.omit(new_df)
#all good now!
```

```{r}
#sex2
new_df <- new_df %>% 
  mutate(sex2 = case_when(sex == 1 ~ "Female",
                          sex == 2 ~ "Male"))
```

```{r}
#age2
new_df <- new_df %>% 
  mutate(age2 = case_when(age == 2 ~ "18-25",
                          age == 3 ~ "26-34",
                          age == 4 ~ "35-49",
                          age == 5 ~ "50-64",
                          age == 6 ~ "65+"))
```

```{r}
#nicdepend2
new_df <- new_df %>% 
  mutate(nicdepend2 = case_when(nicdepend == 0 ~ "No",
                                nicdepend == 1 ~ "Yes"))
```

```{r}
#alcdisorder_year2
new_df <- new_df %>%
  mutate(alcdisorder_year2 = case_when(alcdisorder_year == 0 ~ "No",
                                       alcdisorder_year == 1 ~ "Yes"))
```

```{r}
#mjdisorder_year2
new_df <- new_df %>%
  mutate(mjdisorder_year2 = case_when(mjdisorder_year == 0 ~ "No",
                                      mjdisorder_year == 1 ~ "Yes"))
```

```{r}
#psych_distress_30_2
new_df <- new_df %>% 
  mutate(psych_distress_30_2 = 
           case_when(psych_distress_30 == 0 ~ "No",
                     psych_distress_30 == 1 ~ "Yes"))
```

```{r}
#suicide2
new_df <- new_df %>% 
  mutate(suicide2 = case_when(suicide == 0 ~ "No",
                              suicide == 1 ~ "Yes"))
```

```{r}
#majordepressive_year2
new_df <- new_df %>% 
  mutate(majordepressive_year2 = 
           case_when(majordepressive_year == 0 ~ "No",
                     majordepressive_year == 1 ~ "Yes"))
```

Next, I will change the labels for each of the variables for use later on in tables.

```{r}
new_df = apply_labels(new_df,
                      sex_orient = "Sexual Orientation",
                      sex = "Sex",
                      age = "Age Category",
                      cig30use = "# Days Used Cigarettes in Past Month",
                      nicdepend = "Nicotine Dependence",
                      alc30use = "# Days Drank Alcohol in Past Month",
                      alcbinge30 = "# Days Binge Drank Alcohol in Past Month",
                      mj30use = "# Days Used Marijuana in Past Month",
                      alcdisorder_year = "Alcohol Use Disorder",
                      mjdisorder_year = "Marijuana Use Disorder",
                      psych_distress_30 = "Psychological Distress in Past Month",
                      suicide = "Suicidality in Past Year",
                      majordepressive_year = "Major Depressive Episode in Past Year",
                      sex_orient3 = "Sexual Orientation")
```

#### **Saving Cleaned Data as Parquet File**

There were some issues when I tried saving this dataframe as a CSV, so instead I saved it as a Parquet file. After doing so, I used an online converter in order to convert the Parquet file to a CSV for easier analysis in Tableau later on.

```{r}
write_parquet(new_df, "clean_nsduh_data.parquet")
```

### **Descriptive Statistics**

In this section, I will go over all of the variables of interest relating to this project through a brief description of each variable and the possible values as well as some descriptive statistics.

```{r}
dim(new_df)
```

After being cleaned and processed, the new data frame has 22 columns and 40,389 rows.

1.  **sex_orient**- sexual orientation of respondent-

    -   Values

        -   1 = Heterosexual

        -   2 = Gay/Lesbian

        -   3 = Bisexual

        -   4 = Other

        -   5 = Questioning

```{r}
sex_orient_freq_table <- new_df %>%   
  group_by(sex_orient2) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency)) 

sex_orient_freq_table
```

```{r}
#heterosexual
34587/40389

#bisexual
3020/40389

#gay/lesbian
1212/40389

#other
846/40389

#questioning
724/40389
```

```{r, fig.alt = "A bar plot with sexual orientation (consisting of heterosexual, bisexual, gay/lesbian, other, and questioning) on the x-axis and count on the y-axis. Plot illustrates the distribution of the data in terms of sexual orientation."}
ggplot(new_df, aes(x = fct_infreq(sex_orient2))) +
  geom_bar(fill = "#0073C2FF") +
  labs(title = "Sexual Orientation") +
  xlab("Sexual Orientation") +
  ylab("Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5))
```

The most common sexual orientation among respondents was heterosexual (85.6%) followed by bisexual (7.5%) and gay/lesbian (3%).

2.  **sex**- sex at birth of respondent -

    -   Values

        -   1 = Female

        -   2 = Male

```{r}
new_sex_freq_table <- new_df %>%   
  group_by(sex2) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency)) 

new_sex_freq_table
```

```{r}
#male
22223/40389

#female
18166/40389
```

```{r, fig.alt= "A bar plot with sex (as either male or female) on the  x-axis and count on the y-axis. Plot illustrates the distribution of the data in terms of sex."}
ggplot(new_df, aes(x = fct_infreq(sex2))) +
  geom_bar(fill = "#0073C2FF") +
  labs(title = "Sex at Birth") +
  xlab("Sex") +
  ylab("Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5))
```

55% of respondents were born male, while 45% of respondents were born female.

3.  **age**- age category -

    -   Values

        -   2 = 18-25 years old

        -   3 = 26-34 years old

        -   4 = 35-49 years old

        -   5 = 50-64 years old

        -   6 = 65 or older

```{r}
age_freq_table <- new_df %>%   
  group_by(age2) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))  

age_freq_table
```

```{r}
#18-25
11980/40389

#35-49
11135/40389

#26-34
8234/40389

#65+
4571/40389

#50-64
4469/40389
```

```{r, fig.alt= "A bar plot with age category (in years) on the x-axis and count on the y-axis. Plot illustrates the distribution of the data in terms of age group."}
ggplot(new_df, aes(x = fct_infreq(age2))) +
  geom_bar(fill = "#0073C2FF") +
  labs(title = "Age Category") +
  xlab("Age Category") +
  ylab("Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5)) 
```

The majority of respondents fell into the 18-25 (29.7%) and 35-49 (27.6%) age categories. People who were 26-34 made up 20.4% of the data. The least represented age groups in this data frame are from the 50-64 and 65+ age groups, making up 11.1% and 11.3% respectively.

4.  **cig30use**- during the past 30 days, on how many days did you smoke part of or all of a cigarette? -

    -   Values

        -   0-30

```{r}
new_df$cig30use %>%
  summary()
```

```{r}
cig30use_freq_table <- new_df %>%   
  group_by(cig30use) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))  

cig30use_freq_table
```

```{r}
34679/40389

3104/40389
```

```{r, fig.alt= "A histogram with the number of days respondents smoked cigarettes in the past month on the x-axis and frequency on the y-axis. Plot illustrates the distribution of the data in terms of frequency of smoking cigarettes."}
hist(new_df$cig30use, xlab = "# of Days", main = "# of Days Respondent Smoked Cigarettes in Past Month", col = "#0073C2FF")
```

The mean number of days respondents smoked cigarettes in the past month was 2.9. The majority of respondents said they had not smoked any cigarettes in the past month (85.9%), while the second most common response was 30 days (7.7%).

5.  **nicdepend**- nicotine dependence in past month based on criteria according to the Nicotine Dependence Syndrome Scale and the Fagerstrom Test of Nicotine Dependence -

    -   Values

        -   0 = No

        -   1 = Yes

```{r}
nicdepend_freq_table <- new_df %>%   
  group_by(nicdepend2) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency)) 

nicdepend_freq_table
```

```{r}
#no
37177/40389

#yes
3212/40389
```

```{r, fig.alt = "A bar plot with nicotine dependence on the x-axis (measured as yes/no) and count on the y-axis. Plot illustrates the distribution of the data in terms of whether or not respondents have a dependence on nicotine."}
ggplot(new_df, aes(x = nicdepend2)) +
  geom_bar(fill = "#0073C2FF") +
  labs(title = "Nicotine Dependence") +
  xlab("Nicotine Dependence") +
  ylab("Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5))
```

92% of respondents were not classified as having a dependence on nicotine, compared to 8% of respondents who were classified as having a dependence on nicotine.

6.  **alc30use**- number of days respondent had at least one drink of alcohol in the past 30 days -

    -   Values

        -   0-30

```{r}
new_df$alc30use %>%
  summary()
```

```{r}
alc30use_freq_table <- new_df %>%   
  group_by(alc30use) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))  

alc30use_freq_table
```

```{r}
18851/40389
```

```{r, fig.alt= "A histogram with the number of days respondents had at least one drink of alcohol in the past month on the x-axis and frequency on the y-axis. Plot illustrates the distribution of the data in terms of frequency of drinking at least one drink in the past month."}
hist(new_df$alc30use, xlab = "# of Days", main = "# of Days Respondent Had 1+ Drink of Alcohol", col = "#0073C2FF")
```

The mean number of days respondents had at least one in the past month was 4.1. The most frequent response was that respondents had not had any drinks in the past month (46.7%).

7.  **alcbinge30**- number of days respondent had 4 or more drinks in the past 30 days -

    -   Values

        -   0-30

```{r}
new_df$alcbinge30 %>%
  summary()
```

```{r}
alcbinge30_freq_table <- new_df %>%   
  group_by(alcbinge30) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency)) 

alcbinge30_freq_table
```

```{r}
29965/40389
```

```{r, fig.alt= "A histogram with the number of days respondents had at least four drinks of alcohol in the past month on the x-axis and frequency on the y-axis. Plot illustrates the distribution of the data in terms of frequency of binge drinking in the past month."}
hist(new_df$alcbinge30, xlab = "# of Days", main = "Number of Days With 4+ Drinks", col = "#0073C2FF")
```

The mean number of days a respondent had 4 or more drinks in the past 30 days was 1.1. The most frequent response from respondents was that they had never drank 4 or more drinks in the past month at 74.2%.

8.  **mj30use**- number of days respondent used marijuana/cannabis in any form in the past 30 days - Values

    -   0-30

```{r}
new_df$mj30use %>%
  summary()
```

```{r}
mj30use_freq_table <- new_df %>%   
  group_by(mj30use) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency)) 

mj30use_freq_table
```

```{r}
32344/40389
```

```{r, fig.alt= "A histogram with the number of days respondents used marijuana in the past month on the x-axis and frequency on the y-axis. Plot illustrates the distribution of the data in terms of frequency of marijuana use."}
hist(new_df$mj30use, xlab = "# of Days", main = "Number of Days Used Marijuana", col = "#0073C2FF")
```

The mean number of days a respondent used marijuana in the past 30 days was around 3.3. The most frequent response from respondents was that they had never used marijuana in the past month at 80.1%.

9.  **alcdisorder_year**- whether respondents fits DSM-5 criteria for alcohol use disorder - Values

    -   0 = No

    -   1 = Yes

```{r}
alcdisorder_year_freq_table <- new_df %>%   
  group_by(alcdisorder_year2) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency)) 

alcdisorder_year_freq_table
```

```{r}
#no
35277/40389

#yes
5112/40389
```

```{r, fig.alt= "A bar plot with alcohol use disorder on the x-axis (measured as yes/no) and count on the y-axis. Plot illustrates the distribution of the data in terms of whether or not respondents meet the criteria for alcohol use disorder."}
ggplot(new_df, aes(alcdisorder_year2)) +
  geom_bar(fill = "#0073C2FF") +
  labs(title = "Alcohol Use Disorder in Past Year") +
  xlab("Alcohol Use Disorder") +
  ylab("Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5))
```

87.3% of respondents did not meet the DSM-5 criteria for having alcohol use disorder in the past year. Only 12.7% of respondents did meet the criteria for alcohol use disorder.

10. **mjdisorder_year**- whether respondent fits criteria for marijuana use disorder - Values

    -   0 = No

    -   1 = Yes

```{r}
mjdisorder_year_freq_table <- new_df %>%   
  group_by(mjdisorder_year2) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency)) 

mjdisorder_year_freq_table
```

```{r}
#no
36500/40389

#yes
3889/40389
```

```{r, fig.alt= "A bar plot with marijuana use disorder on the x-axis (measured as yes/no) and count on the y-axis. Plot illustrates the distribution of the data in terms of whether or not respondents meet the criteria for marijuana use disorder."}
ggplot(new_df, aes(mjdisorder_year2)) +
  geom_bar(fill = "#0073C2FF") +
  labs(title = "Marijuana Use Disorder in Past Year") +
  xlab("Marijuana Use Disorder") +
  ylab("Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5))
```

90.4% of respondents did not meet the DSM-5 criteria for having marijuana use disorder in the past year. Only 9.6% of respondents did meet the criteria for marijuana use disorder.

11. **psych_distress_30**- past month serious psychological distress indicator (recoded based on answers to K6 questionnaire) - Values

    -   0 = No

    -   1 = Yes

```{r}
psych_distress_30_freq_table <- new_df %>%   
  group_by(psych_distress_30_2) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))

psych_distress_30_freq_table
```

```{r}
#no
35973/40389

#yes
4416/40389
```

```{r, fig.alt = "A bar plot with serious psychological distress in the past 30 days on the x-axis (measured as yes/no) and count on the y-axis. Plot illustrates the distribution of the data in terms of whether or not respondents experienced serious psychological distress in the past 30 days."}
ggplot(new_df, aes(psych_distress_30_2)) +
  geom_bar(fill = "#0073C2FF") +
  labs(title = "Serious Psychological Distress in Past 30 Days") +
  xlab("Serious Psychological Distress") +
  ylab("Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5))
```

89.1% of respondents have not experienced serious psychological distress in the past 30 days, while 10.9% have.

12. **suicide**- recoded adult past year thoughts, plans, OR attempts of suicide - Values

    -   0 = No

    -   1 = Yes, one or more (thoughts, plans, attempts, or some combination)

```{r}
suicide_freq_table <- new_df %>%   
  group_by(suicide2) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))

suicide_freq_table
```

```{r}
#no
37388/40389

#yes
3001/40389
```

```{r, fig.alt = "A bar plot with suicidal thoughts, plans, and/or attempts in the past year on the x-axis (measured as yes/no)and count on the y-axis. Plot illustrates the distribution of the data in terms of whether or not respondents experienced at least one of the following- suicidal thoughts, plans, and/or attempts."}
ggplot(new_df, aes(suicide2)) +
  geom_bar(fill = "#0073C2FF") +
  labs(title = "Suicidal Thoughts, Plans, And/Or Attempts in Past Year") +
  xlab("Suicidal Thoughts, Plans, And/Or Attempts") +
  ylab("Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5))
```

92.6% of respondents did not have suicidal thoughts, plans, and/or attempts in the past 30 days while 7.4% experienced at least one of these.

13. **majordepressive_year**- past year major depressive episode - Values

    -   0 = No

    -   1 = Yes

```{r}
majordepressive_year_freq_table <- new_df %>%   
  group_by(majordepressive_year2) %>%   
  summarise(frequency = n()) %>%   arrange(desc(frequency))

majordepressive_year_freq_table
```

```{r}
#no
35623/40389

#yes
4766/40389
```

```{r, fig.alt = "A bar plot with major depressive episode in the past year on the x-axis (measured as yes/no) and count on the y-axis. Plot illustrates the distribution of the data in terms of whether or not respondents experienced a major depressive episode in the past year."}
ggplot(new_df, aes(majordepressive_year2)) +
  geom_bar(fill = "#0073C2FF") +
  labs(title = "Experienced Major Depressive Episode in Past Year") +
  xlab("Major Depressive Episode") +
  ylab("Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5))
```

88.2% of respondents did not experience a major depressive episode in the past year, while 11.8% did experience a major depressive episode.

## **Exploring Relationships Within the Data**

### **Nicotine Dependence by Sex, Age, and Sexual Orientation**

#### **Sex and Nicotine Dependence**

```{r, fig.alt = "A bar plot with nicotine dependence (filtering by responses with yes) on the x-axis and count on the y-axis. There are two bars that represent the sex variable (measured as either male or female). Plot illustrates the differences in sex among those with a nicotine dependence."}
new_df %>%
  filter(nicdepend2 %in% c("Yes")) %>%
  ggplot(aes(nicdepend2, fill = sex2)) +
    geom_bar(position = "dodge",
             alpha = 1) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Nicotine Dependence by Sex",
       x = "Nicotine Dependence",
       y = "Count",
       fill = "Sex")
```

Women were slightly more likely to be dependent on nicotine than men.

#### **Age and Nicotine Dependence**

```{r, fig.alt= "A bar plot with nicotine dependence (filtering by responses with yes) on the x-axis and count on the y-axis. There are five bars that represent the age variable (measured as 18-25, 26-34, 35-49, 50-64, and 65+ years old). Plot illustrates the differences in age among those with a nicotine dependence."}
new_df %>%
  filter(nicdepend2 %in% c("Yes")) %>%
  ggplot(aes(nicdepend2, fill = age2)) +
    geom_bar(position = "dodge",
             alpha = 1) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Nicotine Dependence by Age",
       x = "Nicotine Dependence",
       y = "Count",
       fill = "Age")
```

People aged 35-49 were the most likely age group to have a nicotine dependence.

#### **Sexual Orientation and Nicotine Dependence**

```{r, fig.alt = "A bar plot with nicotine dependence (filtering by responses with yes) on the x-axis and count on the y-axis. There are five bars that represent the sexual orientation variable (measured as heterosexual, bisexual, gay/lesbian, other, and questioning). Plot illustrates the differences in sexual orientation among those with a nicotine dependence."}
new_df %>%
  filter(nicdepend2 %in% c("Yes")) %>%
  ggplot(aes(nicdepend2, fill = sex_orient2)) +
    geom_bar(position = "dodge",
             alpha = 1) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Nicotine Dependence by Sexual Orientation",
       x = "Nicotine Dependence",
       y = "Count",
       fill = "Sexual Orientation")
```

Heterosexual people were the most likely sexual orientation group to have a dependence on nicotine, followed by bisexual people.

### **Alcohol Use Disorder by Sex, Age, and Sexual Orientation**

#### **Sex and Alcohol Use Disorder**

```{r, fig.alt = "A bar plot with alcohol use disorder (filtering by responses with yes) on the x-axis and count on the y-axis. There are two bars that represent the sex variable (measured as either male or female). Plot illustrates the differences in sex among those with an alcohol use disorder."}
new_df %>%
  filter(alcdisorder_year2 %in% c("Yes")) %>%
  ggplot(aes(alcdisorder_year2, fill = sex2)) +
    geom_bar(position = "dodge",
             alpha = 1) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Alcohol Use Disorders by Sex",
       x = "Alcohol Use Disorder",
       y = "Count",
       fill = "Sex")
```

Women were slightly more likely to have an alcohol use disorder than men.

#### **Age and Alcohol Use Disorder**

```{r, fig.alt = "A bar plot with alcohol use disorder (filtering by responses with yes) on the x-axis and count on the y-axis. There are five bars that represent the age variable (measured as 18-25, 26-34, 35-49, 50-64, and 65+ years old). Plot illustrates the differences in age among those with an alcohol use disorder."}
new_df %>%
  filter(alcdisorder_year2 %in% c("Yes")) %>%
  ggplot(aes(alcdisorder_year2, fill = age2)) +
    geom_bar(position = "dodge",
             alpha = 1) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Alcohol Use Disorders by Age",
       x = "Alcohol Use Disorder",
       y = "Count",
       fill = "Age")
```

The 18-25 age category had the largest amount of people with an alcohol use disorder, followed by the 35-49 and 26-34 age categories.

#### **Sexual Orientation and Alcohol Use Disorder**

```{r, fig.alt = "A bar plot with alcohol use disorder (filtering by responses with yes) on the x-axis and count on the y-axis. There are five bars that represent the sexual orientation variable (measured as heterosexual, bisexual, gay/lesbian, other, and questioning). Plot illustrates the differences in sexual orientation among those with an alcohol use disorder."}
new_df %>%
  filter(alcdisorder_year2 %in% c("Yes")) %>%
  ggplot(aes(alcdisorder_year2, fill = sex_orient2)) +
    geom_bar(position = "dodge",
             alpha = 1) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Alcohol Use Disorders by Sexual Orientation",
       x = "Alcohol Use Disorder",
       y = "Count",
       fill = "Sexual Orientation")
```

Heterosexual people were the most likely sexual orientation group to have an alcohol use disorder, followed by bisexual people.

### **Marijuana Use Disorder by Sex, Age, and Sexual Orientation**

#### **Sex and Marijuana Use Disorder**

```{r, fig.alt = "A bar plot with marijuana use disorder (filtering by responses with yes) on the x-axis and count on the y-axis. There are two bars that represent the sex variable (measured as either male or female). Plot illustrates the differences in sex among those with a marijuana use disorder."}
new_df %>%
  filter(mjdisorder_year2 %in% c("Yes")) %>%
  ggplot(aes(mjdisorder_year2, fill = sex2)) +
    geom_bar(position = "dodge",
             alpha = 1) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Marijuana Use Disorders by Sex",
       x = "Marijuana Use Disorder",
       y = "Count",
       fill = "Sex")
```

Women were slightly more likely to have a marijuana use disorder than men.

#### **Age and Marijuana Use Disorder**

```{r, fig.alt = "A bar plot with marijuana use disorder (filtering by responses with yes) on the x-axis and count on the y-axis. There are five bars that represent the age variable (measured as 18-25, 26-34, 35-49, 50-64, and 65+ years old). Plot illustrates the differences in age among those with a marijuana use disorder."}
new_df %>%
  filter(mjdisorder_year2 %in% c("Yes")) %>%
  ggplot(aes(mjdisorder_year2, fill = age2)) +
    geom_bar(position = "dodge",
             alpha = 1) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Marijuana Use Disorders by Age",
       x = "Marijuana Use Disorder",
       y = "Count",
       fill = "Age")
```

The 18-25 age category had the largest amount of people with a marijuana use disorder, followed by the 26-34 and 35-49 age categories.

#### **Sexual Orientation and Marijuana Use Disorder**

```{r, fig.alt = "A bar plot with marijuana use disorder (filtering by responses with yes) on the x-axis and count on the y-axis. There are five bars that represent the sexual orientation variable (measured as heterosexual, bisexual, gay/lesbian, other, and questioning). Plot illustrates the differences in sexual orientation among those with a marijuana use disorder."}
new_df %>%
  filter(mjdisorder_year2 %in% c("Yes")) %>%
  ggplot(aes(mjdisorder_year2, fill = sex_orient2)) +
    geom_bar(position = "dodge",
             alpha = 1) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Marijuana Use Disorders by Sexual Orientation",
       x = "Marijuana Use Disorder",
       y = "Count",
       fill = "Sexual Orientation")
```

Heterosexual people were the most likely sexual orientation group to have a marijuana use disorder, followed by bisexual people.

### **Psychological Distress by Sex, Age, and Sexual Orientation**

#### **Sex and Psychological Distress**

```{r, fig.alt = "A bar plot with psychological distress in the past month (filtering by responses with yes) on the x-axis and count on the y-axis. There are two bars that represent the sex variable (measured as either male or female). Plot illustrates the differences in sex among those who have experienced severe psychological distress in the past month."}
new_df %>%
  filter(psych_distress_30_2 %in% c("Yes")) %>%
  ggplot(aes(psych_distress_30_2, fill = sex2)) +
    geom_bar(position = "dodge",
             alpha = 1) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Psychological Distress in Past Month by Sex",
       x = "Psychological Distress in Past Month",
       y = "Count",
       fill = "Sex")
```

Men were much more likely than women to experience severe psychological distress in the past month.

#### **Age and Psychological Distress**

```{r, fig.alt = "A bar plot with psychological distress in the past month (filtering by responses with yes) on the x-axis and count on the y-axis. There are five bars that represent the age variable (measured as 18-25, 26-34, 35-49, 50-64, and 65+ years old). Plot illustrates the differences in age among those who have experienced severe psychological distress in the past month."}
new_df %>%
  filter(psych_distress_30_2 %in% c("Yes")) %>%
  ggplot(aes(psych_distress_30_2, fill = age2)) +
    geom_bar(position = "dodge",
             alpha = 1) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Psychological Distress in Past Month by Age",
       x = "Psychological Distress in Past Month",
       y = "Count",
       fill = "Age")
```

The 18-25 age category was the most common to have experienced severe psychological distress in the past month, followed by the 26-34 and 35-49 age categories.

#### **Sexual Orientation and Psychological Distress**

```{r, fig.alt= "A bar plot with psychological distress in the past month (filtering by responses with yes) on the x-axis and count on the y-axis. There are five bars that represent the sexual orientation variable (measured as heterosexual, bisexual, gay/lesbian, other, and questioning). Plot illustrates the differences in sexual orientation among those who have experienced severe psychological distress in the past month."}
new_df %>%
  filter(psych_distress_30_2 %in% c("Yes")) %>%
  ggplot(aes(psych_distress_30_2, fill = sex_orient2)) +
    geom_bar(position = "dodge",
             alpha = 1) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Psychological Distress in Past Month by Sexual Orientation",
       x = "Psychological Distress in Past Month",
       y = "Count",
       fill = "Sexual Orientation")
```

Heterosexual people were the most likely sexual orientation group to experience severe psychological distress in the past month, followed by bisexual people.

### **Suicidality by Sex, Age, and Sexual Orientation**

#### **Sex and Suicidality**

```{r, fig.alt = "A bar plot with suicidal thoughts, plans, and/or attempts in the past year (filtering by responses with yes) on the x-axis and count on the y-axis. There are two bars that represent the sex variable (measured as either male or female). Plot illustrates the differences in sex among those who have experienced suicidal thoughts, plans, and/or attempts in the past year."}
new_df %>%
  filter(suicide2 %in% c("Yes")) %>%
  ggplot(aes(suicide2, fill = sex2)) +
    geom_bar(position = "dodge",
             alpha = 1) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Suicidal Thoughts, Plans, and/or Attempts in Past Year by Sex",
       x = "Suicidal Thoughts, Plans, and/or Attempts in Past Year",
       y = "Count",
       fill = "Sex")
```

Men were more likely to experience suicidal thoughts, plans, and/or attempts in the past year compared to women.

#### **Age and Suicidality**

```{r, fig.alt = "A bar plot with suicidal thoughts, plans, and/or attempts in the past year (filtering by responses with yes) on the x-axis and count on the y-axis. There are five bars that represent the age variable (measured as 18-25, 26-34, 35-49, 50-64, and 65+ years old). Plot illustrates the differences in age among those who have experienced suicidal thoughts, plans, and/or attempts in the past year."}
new_df %>%
  filter(suicide2 %in% c("Yes")) %>%
  ggplot(aes(suicide2, fill = age2)) +
    geom_bar(position = "dodge",
             alpha = 1) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Suicidal Thoughts, Plans, and/or Attempts in Past Year by Age",
       x = "Suicidal Thoughts, Plans, and/or Attempts in Past Year",
       y = "Count",
       fill = "Age")
```

The 18-25 age category was the most common to have experienced suicidal thoughts, plans and/or attempts in the past year, followed by the 26-34 and 35-49 age categories.

#### **Sexual Orientation and Suicidality**

```{r, fig.alt = "A bar plot with suicidal thoughts, plans, and/or attempts in the past year (filtering by responses with yes) on the x-axis and count on the y-axis. There are five bars that represent the sexual orientation variable (measured as heterosexual, bisexual, gay/lesbian, other, and questioning). Plot illustrates the differences in sexual orientation among those who have experienced suicidal thoughts, plans, and/or attempts in the past year."}
new_df %>%
  filter(suicide2 %in% c("Yes")) %>%
  ggplot(aes(suicide2, fill = sex_orient2)) +
    geom_bar(position = "dodge",
             alpha = 1) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Suicidal Thoughts, Plans, Attempts in Past Year by Sexual Orientation",
       x = "Suicidal Thoughts, Plans, and/or Attempts in Past Year",
       y = "Count",
       fill = "Sexual Orientation")
```

Heterosexual people were the most likely sexual orientation group to have experienced suicidal thoughts, plans and/or attempts in the past year, followed by bisexual people.

### **Major Depressive Episode by Sex, Age, and Sexual Orientation**

#### **Sex and Major Depressive Episode**

```{r, fig.alt = "A bar plot with major depressive episode in the past year (filtering by responses with yes) on the x-axis and count on the y-axis. There are two bars that represent the sex variable (measured as either male or female). Plot illustrates the differences in sex among those who have experienced a major depressive episode in the past year."}
new_df %>%
  filter(majordepressive_year2 %in% c("Yes")) %>%
  ggplot(aes(majordepressive_year2, fill = sex2)) +
    geom_bar(position = "dodge",
             alpha = 1) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Major Depressive Episode in Past Year by Sex",
       x = "MDE in Past Year",
       y = "Count",
       fill = "Sex")
```

Men were much more likely than women to have experienced a major depressive episode in the past year.

```{r, fig.alt = "A bar plot with major depressive episode in the past year (filtering by responses with yes) on the x-axis and count on the y-axis. There are five bars that represent the age variable (measured as 18-25, 26-34, 35-49, 50-64, and 65+ years old). Plot illustrates the differences in age among those who have experienced a major depressive episode in the past year."}
new_df %>%
  filter(majordepressive_year2 %in% c("Yes")) %>%
  ggplot(aes(majordepressive_year2, fill = age2)) +
    geom_bar(position = "dodge",
             alpha = 1) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Major Depressive Episode in Past Year by Age",
       x = "MDE in Past Year",
       y = "Count",
       fill = "Age")
```

The 18-25 age category was the most common to have experienced a major depressive episode in the past year, followed by the 26-34 and 35-49 age categories.

#### **Sexual Orientation and Major Depressive Episode**

```{r, fig.alt = "A bar plot with major depressive episode in the past year (filtering by responses with yes) on the x-axis and count on the y-axis. There are five bars that represent the sexual orientation variable (measured as heterosexual, bisexual, gay/lesbian, other, and questioning). Plot illustrates the differences in sexual orientation among those who have experienced a major depressive episode in the past year."}
new_df %>%
  filter(majordepressive_year2 %in% c("Yes")) %>%
  ggplot(aes(majordepressive_year2, fill = sex_orient2)) +
    geom_bar(position = "dodge",
             alpha = 1) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Major Depressive Episode in Past Year by Sexual Orientation",
       x = "MDE in Past Year",
       y = "Count",
       fill = "Sexual Orientation")
```

Heterosexual people were the most likely sexual orientation group to have experienced a major depressive episode in the past year, followed by bisexual people.

### **Substance Use Disorders and Suicidality**

#### **Nicotine Dependence and Suicidality**

```{r}
tetrachoric(select(new_df, nicdepend, suicide))
```

There is a very weak positive correlation (r = 0.14) between nicotine dependence and suicidal thoughts, plans, and/or attempts in the past year.

#### **Alcohol Use Disorder and Suicidality**

```{r}
tetrachoric(select(new_df, alcdisorder_year, suicide))
```

There is a weak positive correlation (r = 0.27) between alcohol use disorder and suicidal thoughts, plans, and/or attempts in the past year.

#### **Marijuana Use Disorder and Suicidality**

```{r}
tetrachoric(select(new_df, mjdisorder_year, suicide))
```

There is a weak positive correlation (r = 0.38) between marijuana use disorder and suicidal thoughts, plans, and/or attempts in the past year.

### **Substance Use Disorders and Major Depressive Episode**

#### **Nicotine Dependence and Major Depressive Episode**

```{r}
tetrachoric(select(new_df, nicdepend, majordepressive_year))
```

There is a very weak positive correlation (r = 0.12) between nicotine dependence and having a major depressive episode in the past year.

#### **Alcohol Use Disorder and Major Depressive Episode**

```{r}
tetrachoric(select(new_df, alcdisorder_year, majordepressive_year))
```

There is a weak positive correlation (r = 0.26) between alcohol use disorder and having a major depressive episode in the past year.

#### **Marijuana Use Disorder and Major Depressive Episode**

```{r}
tetrachoric(select(new_df, mjdisorder_year, majordepressive_year))
```

There is a weak positive correlation (r = 0.35) between marijuana use disorder and having a major depressive episode in the past year.

## **Logistic Regression Models**

### **Model 1**

Hypothesis 1: LGB individuals who have experienced psychological distress in the past month are more likely to engage in regular substance use (alcohol, cigarette, and marijuana) compared to non-LGB individuals.

```{r}
#logistic regression model with interaction
mod1 <- glm(psych_distress_30 ~ majordepressive_year + suicide + sex + age + sex_orient3 * (alc30use + cig30use + mj30use),
            data = new_df,
            family = binomial)

summary(mod1)
```

```{r, fig.alt= "Table that shows the logistic regression output for Model 1. Columns in the table are 'characteristics' which corresponds to variable labels, 'OR' which shows the odds ratio, 'SE' which shows the standard error, '95% CI' which shows the 95% confidence interval for the odds ratio, and 'p-value' which includes stars to show what level the results are significant at."}
mod1tbl <- tbl_regression(mod1, exponentiate = TRUE) %>%
  add_significance_stars(pattern = "{p.value}{stars}",
                         hide_ci = FALSE, hide_p = FALSE)

mod1tbl
```

This model shows that individuals who have experienced a major depressive episode in the past year are almost 8 times as likely to have experienced severe psychological distress in the past month compared to individuals without such a history (p \< 0.001). Individuals who have had suicidal thoughts, plans, and/or attempts in the past year are around 5.5 times more likely to have experienced psychological distress in the past month compared to individuals without a history of suicidality (p \< 0.001). Men compared to women are around 1.25 times as likely to have experienced severe psychological distress in the past month (p \< 0.001). Older individuals are less likely to experience psychological distress, with each higher age category being associated with around 31% lower odds of severe psychological distress taking place (p \< 0.001). This model shows that, on average, LGB individuals are almost twice as likely to experience psychological distress compared to non-LGB individuals (p \< 0.001). This aligns with pre-existing research indicating more severe mental health struggles in LGBT populations. Frequent cigarette smoking (p \< 0.001) and marijuana use (p \< 0.001) are positively associated with psychological distress, while frequent alcohol consumption is not significantly associated with psychological distress.

Hypothesis 1 is not supported with these results. The relationships between alcohol use and psychological distress (p \< 0.05), cigarette use and psychological distress (p \< 0.05), and marijuana use and psychological distress (p \< 0.001) are all weaker for LGB individuals compared to non-LGB individuals.

### **Model 2**

Hypothesis 2: LGB individuals who report experiencing a past year major depressive episode will be more likely to engage in regular substance use (alcohol, cigarette, and marijuana) compared to non-LGB individuals.

```{r}
mod2 <- glm(majordepressive_year ~ suicide + psych_distress_30 + age + sex + sex_orient3 * (alc30use + cig30use + mj30use),
            data = new_df,
            family = binomial)

summary(mod2)
```

```{r, fig.alt= "Table that shows the logistic regression output for Model 2. Columns in the table are 'characteristics' which corresponds to variable labels, 'OR' which shows the odds ratio, 'SE' which shows the standard error, '95% CI' which shows the 95% confidence interval for the odds ratio, and 'p-value' which includes stars to show what level the results are significant at."}
mod2tbl <- tbl_regression(mod2, exponentiate = TRUE) %>%
  add_significance_stars(pattern = "{p.value}{stars}",
                         hide_ci = FALSE, hide_p = FALSE)

mod2tbl
```

This model shows that older individuals are less likely to experience major depressive episodes, with each higher age category associated with approximately 13% lower odds of a MDE (p \< 0.001). Men compared to women are around 1.6 times as likely to have experienced a MDE in the past year (p \< 0.001). This model also shows that LGB individuals are around twice as likely to experience a major depressive episode compared to non-LGB individuals (p \< 0.001). The number of days respondents drank alcohol in the past month is not significantly associated with the likelihood of experiencing MDEs. Despite this, frequent cigarette use (p \< 0.05) and marijuana use (p \< 0.001) are both associated with a greater likelihood of having experienced a MDE in the past year.

Hypothesis 2 is not supported by these results. There is no significant difference between LGB and non-LGB individuals in terms of the relationship between alcohol use and a MDE. Cigarette use (p \< 0.05) and marijuana use (p \< 0.05) are also more weakly associated with MDEs for LGB individuals compared to non-LGB individuals.

### **Model 3**

Hypothesis 3: LGB individuals who have experienced serious psychological distress in the past month will be more likely to have a higher frequency of binge drinking in the past month compared to non-LGB individuals.

```{r}
mod3 <- glm(psych_distress_30 ~ suicide + majordepressive_year + age + sex + (sex_orient3 * alcbinge30),
            data = new_df,
            family = binomial)

summary(mod3)
```

```{r, fig.alt= "Table that shows the logistic regression output for Model 3. Columns in the table are 'characteristics' which corresponds to variable labels, 'OR' which shows the odds ratio, 'SE' which shows the standard error, '95% CI' which shows the 95% confidence interval for the odds ratio, and 'p-value' which includes stars to show what level the results are significant at."}
mod3tbl <- tbl_regression(mod3, exponentiate = TRUE) %>%
  add_significance_stars(pattern = "{p.value}{stars}",
                         hide_ci = FALSE, hide_p = FALSE)

mod3tbl
```

This model shows that more frequent binge drinking is associated with a higher likelihood of psychological distress (p \< 0.001).

Hypothesis 3 is not supported by this model. The relationship between binge drinking and psychological distress does not significantly differ between LGB and non-LGB individuals.

### **Model 4**

Hypothesis 4: LGB individuals who have had suicidal thoughts, plans, or attempts in the past year will be more likely to have a substance use disorder compared to non-LGB individuals.

```{r}
mod4 <- glm(suicide ~ psych_distress_30 + majordepressive_year + age + sex + (sex_orient3 * (alcdisorder_year + mjdisorder_year + nicdepend)),
            data = new_df,
            family = binomial)

summary(mod4)
```

```{r, fig.alt= "Table that shows the logistic regression output for Model 4. Columns in the table are 'characteristics' which corresponds to variable labels, 'OR' which shows the odds ratio, 'SE' which shows the standard error, '95% CI' which shows the 95% confidence interval for the odds ratio, and 'p-value' which includes stars to show what level the results are significant at."}
mod4tbl <- tbl_regression(mod4, exponentiate = TRUE) %>%
  add_significance_stars(pattern = "{p.value}{stars}",
                         hide_ci = FALSE, hide_p = FALSE)

mod4tbl
```

This model shows that older individuals are significantly less likely to experience suicidal thoughts, plans or attempts, with each higher age category decreasing the risk of suicidality by around 21% (p \< 0.001). Women are around 20% more likely to experience suicidal thoughts, plans or attempts than men (p \< 0.001). This model also shows that LGB individuals are approximately twice as likely to experience suicidal thoughts, plans or attempts than non-LGB individuals (p \< 0.001). Individuals who have either an alcohol use disorder (p \< 0.001) or a marijuana use disorder (p \< 0.001) are around 1.5 times as likely to also experience suicidal thoughts, plans or attempts. Nicotine dependence is not significantly associated with the likelihood of suicidality in this model.

Hypothesis 4 is not supported by this model. Individuals with substance use disorders do not significantly differ in their suicidality based on sexual orientation.

## **Diagnostics/Model Evaluation**

Since binary variables are the only types of predictors utilized in these models, we do not need to check for a linearity assumption.

### **Model 1**

#### **Multicollinearity**

```{r}
#check multicollinearity with variance inflation factors (VIF)- should be below 5
vif(mod1)
```

Since the Variance Inflation Factor is between 1 and 5 for each of the predictor variables, there is no multicollinearity detected in Model 1.

#### **Pseudo R-Squared**

```{r}
#McFadden's R^2 - values closer to 1 mean better predictive power
pR2(mod1)
```

A McFadden’s R-squared value of 0.2967 indicates that Model 1 explains about 29.67% of the variation in the outcome variable.

#### **Accuracy**

```{r}
#predict probabilities
pred_probs1 <- predict(mod1, 
                       newdf1 = select(new_df, -psych_distress_30),
                       type = "response")

#convert probabilities to binary outcomes
pred_outcome1 <- ifelse(pred_probs1 > 0.5, 1, 0)

#transform predictions into factor
pred_outcome1 <- factor(pred_outcome1,
                        levels = c(0, 1))

#create confusion matrix
tab1 = table(new_df$psych_distress_30, pred_outcome1,
            dnn = c("observed", "predicted"))

tab1
```

```{r}
#calculate accuracy
(35171 + 1519) / (802 + 2897 + 35171 + 1519)
```

Model 1 has a predictive accuracy of 90.8%.

#### **Sensitivity and Specificity**

```{r}
#sensitivity
sensitivity1 <- tab1[2, 2] / (tab1[2, 2] + tab1[2, 1])

sensitivity1
```

```{r}
#specificity
specificity1 <- tab1[1, 1] / (tab1[1, 1] + tab1[1, 2])

specificity1
```

Model 1 has a higher specificity (97.8%) than sensitivity (34.4%), which means that it is good at identifying true negatives (those who have not experienced psychological distress) but it might misclassify true positives (those who have experienced psychological distress).

#### **AUC/ROC Curve**

```{r, fig.alt = "A Receiver Operating Characteristic (ROC) curve graph, showing the performance of Model 1 across different thresholds, with the Area Under the Curve (AUC) as a summary metric."}
#the closer the AUC to 1, the better the model fits!
roc1 <- roc(psych_distress_30 ~ fitted(mod1),
            data = new_df)

ggroc(roc1, legacy.axes = TRUE) +
  labs(title = paste0("AUC = ", round(roc1$auc, 2))) +
  theme_minimal() +
  theme(
    panel.grid.major = element_line(color = "gray", size = 0.5),
    panel.grid.minor = element_line(color = "gray", size = 0.25)
  )
```

Model 1 has an AUC of 0.86, which indicates that it is fairly reliable and is performing very well in terms of distinguishing between those who have and have not experienced psychological distress in the past month.

### **Model 2**

#### **Multicollinearity**

```{r}
#check multicollinearity with variance inflation factors (VIF)- should be below 5
vif(mod2)
```

Since the Variance Inflation Factor is between 1 and 5 for each of the predictor variables, there is no multicollinearity detected in Model 2.

#### **Pseudo R-Squared**

```{r}
#McFadden's R^2 - values closer to 1 mean better predictive power
pR2(mod2)
```

A McFadden’s R-squared value of 0.2704 indicates that Model 2 explains about 27.04% of the variation in the outcome variable.

#### **Accuracy**

```{r}
#predict probabilities
pred_probs2 <- predict(mod2, 
                       newdf2 = select(new_df, -majordepressive_year),
                       type = "response")

#convert probabilities to binary outcomes
pred_outcome2 <- ifelse(pred_probs2 > 0.5, 1, 0)

#transform predictions into factor
pred_outcome2 <- factor(pred_outcome2,
                        levels = c(0, 1))

#create confusion matrix
tab2 = table(new_df$majordepressive_year, pred_outcome2,
            dnn = c("observed", "predicted"))

tab2
```

```{r}
#calculate accuracy
(34865 + 1568) / (34865 + 3198 + 758 + 1568)
```

Model 2 has a predictive accuracy of 90.2%.

#### **Sensitivity and Specificity**

```{r}
#sensitivity
sensitivity2 <- tab2[2, 2] / (tab2[2, 2] + tab2[2, 1])

sensitivity2
```

```{r}
#specificity
specificity2 <- tab2[1, 1] / (tab2[1, 1] + tab2[1, 2])

specificity2
```

Model 2 has a higher specificity (97.9%) than sensitivity (32.9%), which means that it is good at identifying true negatives (those who have not experienced a major depressive episode) but it might misclassify true positives (those who have experienced a major depressive episode).

#### **AUC/ROC Curve**

```{r, fig.alt = "A Receiver Operating Characteristic (ROC) curve graph, showing the performance of Model 2 across different thresholds, with the Area Under the Curve (AUC) as a summary metric."}
#the closer the AUC to 1, the better the model fits!
roc2 <- roc(majordepressive_year ~ fitted(mod2),
            data = new_df)

ggroc(roc2, legacy.axes = TRUE) +
  labs(title = paste0("AUC = ", round(roc2$auc, 2))) +
  theme_minimal() +
  theme(
    panel.grid.major = element_line(color = "gray", size = 0.5),
    panel.grid.minor = element_line(color = "gray", size = 0.25)
  )
```

Model 2 has an AUC of 0.83, which indicates that it is fairly reliable and is performing very well in terms of distinguishing between those who have and have not experienced a major depressive episode in the past year.

### **Model 3**

#### **Multicollinearity**

```{r}
#check multicollinearity with variance inflation factors (VIF)- should be below 5
vif(mod3)
```

Since the Variance Inflation Factor is between 1 and 5 for each of the predictor variables, there is no multicollinearity detected in Model 3.

#### **Pseudo R-Squared**

```{r}
#McFadden's R^2 - values closer to 1 mean better predictive power
pR2(mod3)
```

A McFadden’s R-squared value of 0.2855 indicates that Model 3 explains about 28.55% of the variation in the outcome variable.

#### **Accuracy**

```{r}
#predict probabilities
pred_probs3 <- predict(mod3, 
                       newdf3 = select(new_df, -psych_distress_30),
                       type = "response")

#convert probabilities to binary outcomes
pred_outcome3 <- ifelse(pred_probs3 > 0.5, 1, 0)

#transform predictions into factor
pred_outcome3 <- factor(pred_outcome3,
                        levels = c(0, 1))

#create confusion matrix
tab3 = table(new_df$psych_distress_30, pred_outcome3,
            dnn = c("observed", "predicted"))

tab3
```

```{r}
#calculate accuracy
(35216 + 1445) / (35216 + 2971 + 757 + 1445)
```

Model 3 has a predictive accuracy of 90.8%.

#### **Sensitivity and Specificity**

```{r}
#sensitivity
sensitivity3 <- tab3[2, 2] / (tab3[2, 2] + tab3[2, 1])

sensitivity3
```

```{r}
#specificity
specificity3 <- tab3[1, 1] / (tab3[1, 1] + tab3[1, 2])

specificity3
```

Model 3 has a higher specificity (97.9%) than sensitivity (32.7%), which means that it is good at identifying true negatives (those who have not experienced psychological distress) but it might misclassify true positives (those who have experienced psychological distress).

#### **AUC/ROC Curve**

```{r, fig.alt = "A Receiver Operating Characteristic (ROC) curve graph, showing the performance of Model 3 across different thresholds, with the Area Under the Curve (AUC) as a summary metric."}
#the closer the AUC to 1, the better the model fits!
roc3 <- roc(psych_distress_30 ~ fitted(mod3),
            data = new_df)

ggroc(roc3, legacy.axes = TRUE) +
  labs(title = paste0("AUC = ", round(roc3$auc, 2))) +
  theme_minimal() +
  theme(
    panel.grid.major = element_line(color = "gray", size = 0.5),
    panel.grid.minor = element_line(color = "gray", size = 0.25)
  )
```

Model 3 has an AUC of 0.85, which indicates that it is fairly reliable and is performing very well in terms of distinguishing between those who have and have not experienced psychological distress in the past month.

### **Model 4**

#### **Multicollinearity**

```{r}
#check multicollinearity with variance inflation factors (VIF)- should be below 5
vif(mod4)
```

Since the Variance Inflation Factor is between 1 and 5 for each of the predictor variables, there is no multicollinearity detected in Model 4.

#### **Pseudo R-Squared**

```{r}
#McFadden's R^2 - values closer to 1 mean better predictive power
pR2(mod4)
```

A McFadden’s R-squared value of 0.3011 indicates that Model 4 explains about 30.11% of the variation in the outcome variable.

#### **Accuracy**

```{r}
#predict probabilities
pred_probs4 <- predict(mod4, 
                       newdf4 = select(new_df, -suicide),
                       type = "response")

#convert probabilities to binary outcomes
pred_outcome4 <- ifelse(pred_probs4 > 0.5, 1, 0)

#transform predictions into factor
pred_outcome4 <- factor(pred_outcome4,
                        levels = c(0, 1))

#create confusion matrix
tab4 = table(new_df$suicide, pred_outcome4,
            dnn = c("observed", "predicted"))

tab4
```

```{r}
#calculate accuracy
(36683 + 904) / (36683 + 2097 + 705 + 904)
```

Model 4 has a predictive accuracy of 93.1%.

#### **Sensitivity and Specificity**

```{r}
#sensitivity
sensitivity4 <- tab4[2, 2] / (tab4[2, 2] + tab4[2, 1])

sensitivity4
```

```{r}
#specificity
specificity4 <- tab4[1, 1] / (tab4[1, 1] + tab4[1, 2])

specificity4
```

Model 4 has a higher specificity (98.1%) than sensitivity (30.1%), which means that it is good at identifying true negatives (those who have not experienced suicidality) but it might misclassify true positives (those who have experienced suicidality).

#### **AUC/ROC Curve**

```{r, fig.alt = "A Receiver Operating Characteristic (ROC) curve graph, showing the performance of Model 4 across different thresholds, with the Area Under the Curve (AUC) as a summary metric."}
#the closer the AUC to 1, the better the model fits!
roc4 <- roc(suicide ~ fitted(mod4),
            data = new_df)

ggroc(roc4, legacy.axes = TRUE) +
  labs(title = paste0("AUC = ", round(roc4$auc, 2))) +
  theme_minimal() +
  theme(
    panel.grid.major = element_line(color = "gray", size = 0.5),
    panel.grid.minor = element_line(color = "gray", size = 0.25)
  )
```

## **Discussion**

### **Psychological Distress**

Individuals who have experienced a major depressive episode (MDE) in the past year are nearly 8 times more likely to experience severe psychological distress in the past month. Those with suicidal plans, thoughts, or attempts are 5.5 times more likely to experience psychological distress. This is in line with prior research that has shown that major depressive episodes and suicidality can lead to greater psychological distress. Since depression and suicidality are both often characterized by feelings of hopelessness, worthlessness, and extreme sadness, it makes sense that these feelings would contribute to heightened distress. Men were also 1.25 times more likely to experience distress than women in this sample. We also found that older adults are much less likely to experience psychological distress compared to young adults, with each increasing age category being associated with around 31% lower odds of severe psychological distress taking place. Older adults have been shown in much previous research to be less likely to experience negative mental health compared to young adults (SAMHSA, 2021), which is in line with our results. LGB individuals are almost twice as likely to experience psychological distress compared to non-LGB individuals. This makes sense considering that LGB individuals tend to face higher rates of mental health challenges due to factors like social stigma and discrimination (King et al., 2008). Cigarette and marijuana use are both positively associated with psychological distress, but alcohol consumption is not significantly linked. In previous research, cigarette, marijuana, and alcohol use have all been linked to more negative mental health outcomes including psychological distress, major depressive episodes, and suicidality (Currie et al., 2005; Poorolajal & Darvishi, 2016; Shamabadi et al., 2023). While our results regarding cigarette and marijuana use were in line with this pre-existing research, our results actually showed that alcohol seemed to have a slightly protective factor, in that more frequent use was associated with a slightly lower risk of psychological distress.

### **Major Depressive Episodes (MDEs)**

Older individuals are less likely to experience MDEs, with each higher age category reducing the odds by 13%. Men are 1.6 times more likely to experience a MDE than women, and LGB individuals are about twice as likely compared to non-LGB individuals to experience a MDE. Cigarette and marijuana use are linked to an increased likelihood of MDEs, while alcohol consumption does not have a significant association.

### **Suicidality**

Older individuals are less likely to experience suicidality, with each higher age category reducing the risk by 21%. Women in our sample were 20% more likely to experience suicidality than men, which is in line with previous research showing that women are significantly more likely to report a past suicide attempt than men (Kelly et al., 2020). LGB individuals were twice as likely as non-LGB individuals to experience suicidal thoughts, plans, or attempts. People with alcohol and marijuana use disorders are 1.5 times more likely to experience suicidality, but nicotine dependence did not show a significant link.

## **Conclusion**

The hypotheses surrounding the relationship between substance use (alcohol, cigarettes, and marijuana) and mental health outcomes for LGB individuals were not supported. Previous research has shown that there are links between LGBT people and substance use disorders, but that was not the case in this sample. There were some limitations in this study that may have caused this disparity. Since the study predominantly includes non-LGB individuals and LGB individuals only made up around 14.4% of the sample, it may not fully capture the unique mental health challenges faced by sexual minorities. In future research, it might be smart to focus on a sample consisting of equal parts LGBTQ+ and cisgender heterosexual individuals by oversampling LGBTQ+ people. Another issue with this study was that it did not include information on transgender or gender diverse individuals. Future studies should try to include this population in research, since they tend to experience even worse mental health outcomes than the lesbian, gay and bisexual populations due to even higher stigma and discrimination (Hajek et al., 2023).

## **Bibliography**

Centers for Disease Control and Prevention. (2024, October 29). Suicide data and statistics. CDC Website. <https://www.cdc.gov/suicide/facts/data.html>.

Currie, S.R. et al. (2005). Comorbidity of major depression with substance use disorders. *The Canadian Journal of Psychiatry, 50*(10), 660-666. <https://doi.org/10.1177/070674370505001013>.

Hajek, A. et al. (2023). Prevalence and determinants of depressive and anxiety symptoms among transgender people: results of a survey. *Healthcare, 11*(705), 1-12. <https://doi.org/10.3390/healthcare11050705>.

Kelly, L.M., Rash, C.J., Alessi, S.M., & Zajac, K. (2020). Correlates and predictors of suicidal ideation and substance use among adults seeking substance use treatment with varying levels of suicidality. *Journal of Substance Abuse and Addiction Treatment, 119*, 1-29. <https://doi.org/10.1016/j.jsat.2020.108145>.

King, M. et al. (2008). A systematic review of mental disorder, suicide, and deliberate self-harm in lesbian, gay, and bisexual people. *BMC Psychiatry 8*(70), 1-17. <https://doi.org/10.1186/1471-244X-8-70>.

Piscopo, K. & Lipari, R.N. (2016). *Suicidal thoughts and behavior among adults: results from the 2015 national survey on drug use and health*. Substance Abuse and Mental Health Services Administration. <https://www.samhsa.gov/data/sites/default/files/NSDUH-DR-FFR3-2015/NSDUH-DR-FFR3-2015.pdf>.

Poorolajal, J. & Darvishi, N. (2016). Smoking and suicide: a meta-analysis. *PLoS ONE 11*(7), 1-18. <https://doi.org/10.1371/journal.pone.0156348>.

Shamabadi, A. et al. (2023). Suicidality risk after using cannabis and cannabinoids: an umbrella review. *Dialogues in Clinical Neuroscience 25*(1), 50-63. <https://doi.org/10.1080/19585969.2023.2231466>.

Substance Abuse and Mental Health Services Administration. (2021). *Key substance use and mental health indicators in the United States: Results from the 2020 National Survey on Drug Use and Health* (HHS Publication No. PEP21-07-01-003, NSDUH Series H-56). Rockville, MD: Center for Behavioral Health Statistics and Quality, Substance Abuse and Mental Health Services Administration. Retrieved from <https://www.samhsa.gov/data/>.
